Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-08-04 ltx/data_paper/DataPaper.tex
edits to introduction of paper added epilepsy example

The human brain brain is arguably the most complicated system known to man. The development of a complete theory for its function is one of the greatest challenges faced by scientists today. To address this challenge, researchers have naturally used both theoretical and experimental frameworks. Experimental frameworks are typically designed to uncover causal relationships between the systems properties or parameters and its function (reverse engineering the brain). Theoretical studies are typically used in one of two ways. The first is to explain data acquired in an experiment and the second is to predict system behaviour.

@@

Although the explanatory power of the theoretical models is rather vast, a limitation has been establishing a rigorous method for choosing model parameters. For example, over the last decade it has come to light that there is a significant variability in parameters of neural models across subjects (ref Marder 2006). For example, it has been shown that a target network behaviour may be achieved by continuum of parameter combinations (Tail wagging the ref). We expect an analogous situation in disease states and pathological activity such as epileptic seizures. Therefore, for models to be clinically useful they must be subject-specific.

To further motivate the development of a framework for patient-specific neural models we will use intracranial EEG monitoring for planning epilepsy surgery as an example. During iEEG monitoring high fidelity spatiotemporal recordings of the electrical fields of the brain are undertaken. The goal is of the clinical work is to record the electrical activity of seizures and locate their origin. This information along with other behaviour and psychophyics studies enables the definition of surgical margins for the resection of pathological tissue. The ability to fuse the iEEG with a model will not only provide

The aim of this paper is to create a patient-specific neural field model, by estimating parameters of a general mathematical model using data acquired from an epilepsy surgical candidate in a clinical setting.

Instead of extracting features we aim to extract meaningful parameters...

>>>>

The human brain brain is arguably the most complicated system known to man. The development of a complete theory for its function is one of the greatest challenges faced by scientists today. To address this challenge, researchers have naturally used both theoretical and experimental frameworks to develop and test hypotheses investigating brain function. Experimental frameworks are typically designed to uncover causal relationships between the systems properties or parameters and its function (reverse engineering the brain). Theoretical studies are typically used in one of two ways. The first is to explain data acquired in an experiment and the second is to predict system behaviour.

@@

Although the predictive explanatory power of the theoretical models is rather vast when they are tuned to mimic experimental conditions, a limitation has been establishing a rigorous method for choosing model parameters in more general situations. For example, over the last decade it has come to light that there is a significant variability in parameters of neural models across subjects (ref Marder 2006). For example, it has been shown that a target network behaviour may be achieved by continuum of parameter combinations (Tail wagging the ref). We expect an analogous situation in disease states and pathological activity such as epileptic seizures. Therefore, for models to be clinically useful they must be subject-specific.

By assimilating experimental or clinical measurements with mathematical models one can infer unmeasured or latent system properties or parameters. Inference of patient-specific parameters has the the potential to revolutionise the treatment of disease. Typically, one can measure the blood-oxygen level dependent signal through fMRI or the electromagnetic fields of the brain through EEG or MEG. In clinical neurology, these measurement modalities give a kind or a `where and when' indication of the presence of a pathology, but provide little information on the causative mechanisms for disease. A classic example of this is in epilepsy management, where the disease is diagnosed using EEG. If electrical seizures are recorded, a medication plan is prescribed. The choice of medication is not based on the data acquired from the EEG, but based on the clinician's experience with similar cases and sometimes trial and error because similar electrographic seizure can have fundamentally different mechanisms of initiation (JTs paper). Ideally in epilepsy monitoring we would like to image the concentration of ions, the synaptic dynamics, the connectivity strength and structure etc. to better inform a treatment plan by understand physiological changes that lead to seizures. This is also highly desirable for epileptic seizure prediction..

The aim of this paper is to contribute towards to goal of being able to infer patient-specific parameters from electrophysiological data.

Contributions from other groups...
